abxD01 Paper

### 2015-03-07 PAPER.Rmd
Edit random forest paragraph

***Modeling microbiota data can predict C. difficile colonization outcomes.*** Given our ability to make general predictions of *C. difficile* colonization based on the microbiota, we next designed a model to calculate expected colonization levels given the relative abundance of a subset of OTUs. Combining the three datasets (Table 1, Supplemental Table 1 and 2), we set a 1% relative abundance threshold, eliminating OTUs that on average were not present at this level in at least one of the 16 experimental groups. In total, there were 44 OTUs utilized for building a random forest regression model. This model fit the data with an R2 of 0.98 (Figure 5A). We identified OTUs that contributed most to increased accuracy of the model using random forest’s feature selection (Figure 5B). Many of the top contributing OTUs includes members of the *Porphyromonadaceae* family, *Alistipes*, *Lachnospiraceae*, *Lactobacillus*, and *Escherichia* taxa. The relationship between these OTUs and *C. difficile* is shown in Figure 5C. *Escherichia* had the largest positive correlation with *C. difficile* levels of any bacterial species.  This finding was consistent across susceptible antibiotic treated communities except for streptomycin treated communities, which altogether lacked *Enterobacteriaceae* spp. OTUs 3 and 39 (*Porphyromonadaceae* and *Lachnospiraceae*, respectively), were the two most important features in accurately predicting subsequent *C. difficile* levels. In fact, a model built with only these two features was able to account for 79% of the variance observed. Together these results suggest that colonization resistance is likely conferred by *porphyromonadaceae*, *Alistipes*, and *lachnospiraceae* and a loss in these bacteria, concurrently with a gain of *Escherichia*, can result in increased susceptibility to infection.

@@

**Figure 5. Random forest model to predict *C. difficile* colonization levels based on the microbiota.** a) The fit of the random forest model by experimental group. The dashed line represents the y=x line. b) The 15 OTUs with the highest percent increase in mean squared error when dropped from the model. c) The percent relative abundance of the top 8 OTUs from (b) on day 0 are plotted against the log<sub>10</sub> *C. difficile* CFU/g feces.

>>>>

***The composition of the microbiota on day 0 is predictive of C. difficile colonization levels on day 1.*** Given our ability to make qualitative predictions of *C. difficile* colonization based on the microbiota, we next designed a statistical model to calculate expected colonization levels given the relative abundance of a subset of OTUs. We used a regression-based random forest machine learning algorithm to predict the level of *C. difficile* colonization observed in the three sets of experiments based on the composition of the microbiota at the time of challenge. The model explained XXX98.0XXX% of the variation in the observed *C. difficile* colonization levels (Figure 5). Many of the OTUs that contributed the most to the quality of the fit included members of the *Porphyromonadaceae*, *Alistipes*, *Lachnospiraceae*, *Lactobacillus*, and *Escherichia*. These results further validate the observations from the correlation-based analysis. Together these results suggest that colonization resistance is likely conferred by *porphyromonadaceae*, *Alistipes*, and *lachnospiraceae* and a loss in these bacteria, concurrently with a gain of *Escherichia*, can result in increased susceptibility to infection. <needs help>Interestingly, the relationship between these OTUs and colonization resistance was not a simple dichotomous or linear relationship. For example, among the streptomycin-treated mice which altogether lacked the *Enterobacteriaceae*, *Porphyromonadaceae* (OTU 3) and *Lachnospiraceae* (OTU 39), were the two most important features in accurately predicting subsequent *C. difficile* levels. In fact, a model built with only these two features was able to account for 79% of the variance observed. </needs help>

@@

**Figure 5. Random forest model to predict *C. difficile* colonization levels based on the microbiota.** (A) The fit of the random forest model by experimental group. The dashed line represents the y=x line. (B) The 15 OTUs with the highest percent increase in mean squared error when dropped from the model. (C) The percent relative abundance of the top 8 OTUs from (B) on day 0 are plotted against the log<sub>10</sub> *C. difficile* CFU/g feces.
